pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-20 manuscript/greenwell.tex
minor edits

The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options (e.g., the option to add a LOESS smooth). Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.

Currently supported models are described in Table~\ref{tab:models}. In a lot of these cases, \code{partial} should be able to automatically determine \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data). In other situations, the user may need to specify these additional arguments. This allows \code{partial} to be flexible enough to handle many of the model types not listed in in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd)
plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
\strong{Note:} the default color map for contour plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a third degree MARS model. \textit{Left}: Default plot. \textit{Middle}: Using a user-specified color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}

>>>>

The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options. For example, \code{plotPartial} makes it straight forward to add a LOESS smooth, or produce a 3-D surface instead of a false color level plot (the default). Of course, since the default output produced by \code{partial} is still a data frame, the user can easily use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.

Currently supported models are described in Table~\ref{tab:models}. In a lot of these cases, \code{partial} should be able to automatically determine appropriate values for the options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data). In other situations, the user may need to specify one or both of these additional arguments. This allows \code{partial} to be flexible enough to handle many of the model types not listed in in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
rwb <- colorRampPalette(c("red", "white", "blue"))
pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd)
plotPartial(pd, contour = TRUE, col.regions = rwb)
plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a third degree MARS model. \textit{Left}: Default plot. \textit{Middle}: Adds contour lines and uses a different color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}
